---
title: New York Times crossword analysis (revised)
date: '`r format(as.Date("2026-02-05"), "%B %d, %Y")`'
format: 
  html: 
    page-layout: full
---
```{r setup, include = FALSE, warning = FALSE, echo = FALSE}
# Load in libraries ----
library(tidyverse)
library(plotly)
library(scales)
library(MetBrewer)
library(ggridges)

# Read in data file ----
nyt <-
  bind_rows(read.csv("./_data/nyt_data_1.csv"),
            read.csv("./_data/nyt_data_2.csv")) |>
  distinct() |>
  select(-c(editor, format_type, publish_type, title, version)) |>
  filter(!is.na(solving_seconds)) |>
  mutate(
    date = as.Date(print_date),
    year = as.integer(format(date , "%Y")),
    day_of_week_name = fct_reorder(day_of_week_name, day_of_week_integer),
    solving_minutes = solving_seconds / 60,
    finished = ifelse(percent_filled != 100, "Unfinished", "Complete"),
    star = ifelse(star == "", "Not gold", star)
  ) |>
  filter(year != 2026) # NOTE: this year just started

theme_pal <- "Hiroshige" # NOTE: 10

plot_func <- function(df, x, y, fillcolour, title, xlab, ylab) {
  ggplot(df, aes(x = x, y = y)) +
    geom_bar(stat = "identity",
             position = "stack",
             fill = fillcolour) +
    theme_classic() +
    labs(title = title, x = xlab, y = ylab)
}

## Part 1 ----
nyt_mtw <-
  nyt |>
  filter(day_of_week_integer %in% c(1, 2, 3))

## Breakdown by day & year ----
solved_day_year <-
  nyt |>
  filter(solved == "True") |>
  group_by(day_of_week_name, year) |>
  count()

### Breakdown by day ----
solved_day <-
  nyt |>
  filter(solved == "True") |>
  group_by(day_of_week_name) |>
  count()

### Breakdown by year ----
solved_year <-
  nyt |>
  filter(solved == "True") |>
  group_by(year) |>
  count()

## Progress by day & year ----
progress_day_year <-
  nyt |>
  filter(solved == "True") |>
  summarise(avg = round(mean(solving_minutes), 2),
            .by = c("day_of_week_name", "year"))

### Progress by day ----
progress_day <-
  nyt |>
  filter(solved == "True") |>
  summarise(avg = round(mean(solving_minutes), 2), .by = c("day_of_week_name"))

### Progress by year ----
progress_year <-
  nyt |>
  filter(solved == "True") |>
  summarise(avg = round(mean(solving_minutes), 2), .by = c("year"))

## Attempts ----
attempts <-
  nyt |>
  count(year, finished) |>
  group_by(year) |>
  mutate(
    fraction = n / sum(n),
    ymax = cumsum(fraction),
    ymin = c(0, head(ymax, n = -1))
  ) |>
  ungroup()

## Gold ----
gold <-
  nyt |>
  count(year, star) |>
  group_by(year) |>
  mutate(
    fraction = n / sum(n),
    ymax = cumsum(fraction),
    ymin = c(0, head(ymax, n = -1))
  ) |>
  ungroup()

gold_all <-
  nyt |>
  count(year, day_of_week_name, star)

knitr::opts_chunk$set(echo =  FALSE, warning = FALSE, message = FALSE)
```

## Background

My partner and I started doing crosswords sometime in 2017 and we've been hooked since. This led us to getting a New York Times (NYT) crossword subscription in 2018. The NYT app shows you when you complete a crossword with a gold or blue star, where the former means you solved it without revealing any answers **and** before the answers come out. The app also provides *some* stats but I wanted to see what else is the data telling us.

## The data

I found a Python package by [Matt Dodge](https://github.com/mattdodge/nyt-crossword-stats) that retrieves your NYT crossword stats and used this to pull in our stats. This data spans from `r format(as.Date("2018-01-01"), "%B %d, %Y")` to `r format(as.Date("2023-01-23"), "%B %d, %Y")`.

## The analysis

::: {.callout-tip collapse="true"}
## Psst...
I'm using the "Hiroshige" palette from the [`MetBrewer`](https://github.com/BlakeRMills/MetBrewer) package.
:::

::: {.callout-caution}
This is a work in progress.
:::

```{r totalcompletes}
solves <- nyt |> filter(finished=='Complete') |> count()
```

As of December 2025, we've completed `{r} solves` NYT crossword puzzles.

🎉🎉🎉

For the time being, here are my plots so far:

## How are we doing so far?

::: {.panel-tabset}

### All days
```{r alldays}
#| fig-width: 12
ggplot(nyt, aes(x = solving_minutes, y = day_of_week_name, colour=met.brewer(theme_pal)[1])) +
  geom_point(alpha = 0.5) +
  scale_y_discrete(limits = rev) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(title = "How long does it take us to solve a puzzle?", x = "Minutes to solve", y = "Day of the week")
```

### Let's focus on some days
```{r somedays}
#| fig-width: 12
ggplot(nyt_mtw,
       aes(x = solving_minutes, y = day_of_week_name, size = solving_minutes, colour=met.brewer(theme_pal)[2])) +
  geom_point(alpha = 0.25) +
  scale_y_discrete(limits = rev) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(title = "How long does it take us to solve a puzzle?", x = "Minutes to solve", y = "Day of the week")
```
:::

You can see that we're pretty comfortable in the Monday to Wednesday range. Typically, Mondays are the easiest and the difficulty increases as you continue in the week. Depending on who you ask Wednesdays and Thursdays could be a toss-up for difficulty. Wednesdays normally have a theme which I feel helps me solve the puzzle easier than if it didn't have a theme. I also just love a theme. 

## What are we solving here?

Here we're looking at the number of solved crosswords.

::: {.panel-tabset}

### Density plots

::: {.panel-tabset}

### By weekday

```{r solveggridges}
#| fig-width: 12
ggplot({nyt |>
    filter(solved == "True")},
       aes(y = day_of_week_name, x = solving_minutes, fill = day_of_week_name)) +
  geom_density_ridges(alpha = 0.6, stat = "binline", bins = 20) +
  theme_ridges() +
  theme(
    legend.position = "none",
    panel.spacing = unit(0.1, "lines"),
    strip.text.x = element_text(size = 8),
    axis.title.y = element_blank()
  ) +
  scale_fill_manual(values = met.brewer(theme_pal)[1:7]) +
  scale_y_discrete(limits = rev) +
  labs(title = "Time to finish a crossword", x = "Solve time (in minutes)")
```

### By year

```{r solveggridges2}
#| fig-width: 12
ggplot({nyt |>
    filter(solved == "True") |> 
    mutate(year = as.character(year))},
    aes(y = year, x = solving_minutes, fill = year)) +
  geom_density_ridges(alpha = 0.6, stat = "binline", bins = 20) +
  theme_ridges() +
  theme(
    legend.position = "none",
    panel.spacing = unit(0.1, "lines"),
    strip.text.x = element_text(size = 8),
    axis.title.y = element_blank()
  ) +
  scale_fill_manual(values = met.brewer(theme_pal)[1:8]) +
  scale_y_discrete(limits = rev) +
  labs(title = "Time to finish a crossword", x = "Solve time (in minutes)")
```

:::

### Day of the week and year

```{r solveweekyear}
#| fig-width: 12
ggplot(solved_day_year, 
       aes(x = year, 
           y = n)) +
  geom_bar(stat = "identity",
           position = "stack",
           fill = met.brewer(theme_pal)[4]) +
  theme_classic() +
  labs(title = "Distribution of solved crosswords by solve status", 
       x = "Year",
       y = "Count of solved crosswords") +
  facet_wrap(
    vars(day_of_week_name), 
    strip.position = "top") +
  theme(strip.background = element_blank(), strip.placement = "outside")
```

### Day of the week
```{r solveweek}
#| fig-width: 12
plot_func(
  solved_day,
  solved_day$day_of_week_name,
  solved_day$n,
  met.brewer(theme_pal)[5],
  "Distribution of solved crosswords by solve status",
  "Day",
  "Solved crosswords"
)
```

### Year
```{r solveyear}
#| fig-width: 12
plot_func(
  solved_year,
  solved_year$year,
  solved_year$n,
  met.brewer(theme_pal)[6],
  "Distribution of solved crosswords by solve status",
  "Year",
  "Solved crosswords"
)
```

:::

## Gotta go fast

::: {.panel-tabset}

### Day of the week and year

```{r timeweekyear}
#| fig-width: 12
ggplot(progress_day_year, 
       aes(x = year, 
           y = avg)) +
  geom_bar(stat = "identity",
           position = "stack",
           fill = met.brewer(theme_pal)[7]) +
  theme_classic() +
  labs(title = "Average time to finish a crossword", 
       x = "Year",
       y = "Average solving time (in minutes)") +
  facet_wrap(
    vars(day_of_week_name), 
    strip.position = "top") +
  theme(strip.background = element_blank(), strip.placement = "outside")
```

### Day of the week

```{r timeweek}
#| fig-width: 12
plot_func(
  progress_day,
  progress_day$day_of_week_name,
  progress_day$avg,
  met.brewer(theme_pal)[8],
  "Average time to finish a crossword",
  "Day",
  "Average solving time (in minutes)"
)
```

### Year
```{r timeyear}
#| fig-width: 12
plot_func(
  progress_year,
  progress_year$year,
  progress_year$avg,
  met.brewer(theme_pal)[9],
  "Average time to finish a crossword",
  "Year",
  "Average solving time (in minutes)"
)
```

:::

## A for effort

::: {.panel-tabset}

### Donut
```{r attemptsdonut}
#| fig-width: 12
ggplot(attempts,
       aes(
         ymax = ymax,
         ymin = ymin,
         xmax = 4,
         xmin = 3,
         fill = finished
       )) +
  geom_rect() +
  coord_polar(theta = "y") + # Try to remove that to understand how the chart is built initially
  xlim(c(2, 4)) + # Try to remove that to see how to make a pie chart
  theme_void() +
  scale_fill_manual(values = c(met.brewer(theme_pal)[10], met.brewer(theme_pal)[1])) +
  labs(title = "Percentage of crosswords by completion status", fill = "Completion status") +
  facet_wrap( ~ year, nrow = 2, ncol = 4)
```

### Line chart

```{r attemptsline}
#| fig-width: 12
ggplot(attempts, aes(x = year, y = n, colour = finished)) +
  geom_line(size = 2) +
  theme_classic() +
  labs(title = "Count of crosswords attempted by completion status", x = "Year", y = "Number of crosswords", colour ="Completion status") +
  scale_fill_manual(values = c(met.brewer(theme_pal)[9], met.brewer(theme_pal)[2]))
```

:::

## ⭐

::: {.panel-tabset}

### Bar

```{r goldbar}
#| fig-width: 12
ggplot(gold_all, aes(x = year, y = n, fill = star)) +
  geom_bar(stat = "identity", position = position_dodge(preserve = "single")) +
  theme_classic() +
  labs(title = "Count of crosswords attempted by year, weekday, and star status", x = "Year", y = "Number of crosswords", fill = "Star status") +
  scale_fill_manual(values = c(met.brewer(theme_pal)[1], met.brewer(theme_pal)[10])) +
  facet_wrap(
    ~ day_of_week_name,
    strip.position = "top"
  ) +
  theme(strip.background = element_blank(), strip.placement = "outside")
```

### Donut
```{r golddonut}
#| fig-width: 12
ggplot(gold,
       aes(
         ymax = ymax,
         ymin = ymin,
         xmax = 4,
         xmin = 3,
         fill = star
       )) +
  geom_rect() +
  coord_polar(theta = "y") + # Try to remove that to understand how the chart is built initially
  xlim(c(2, 4)) + # Try to remove that to see how to make a pie chart
  theme_void() +
  scale_fill_manual(values = c(met.brewer(theme_pal)[8], met.brewer(theme_pal)[3])) +
  labs(title = "Percentage of crosswords by star status", fill = "Star status") +
  facet_wrap( ~ year, nrow = 2, ncol = 4)
```

### Line chart

```{r goldline}
#| fig-width: 12
ggplot(attempts, aes(x = year, y = n, colour = finished)) +
  geom_line(size = 2) +
  theme_classic() +
  labs(title = "Count of crosswords attempted by star status", x = "Year", y = "Number of crosswords", colour ="Star status") +
  scale_fill_manual(values = c(met.brewer(theme_pal)[7], met.brewer(theme_pal)[4]))
```

:::