---
title: "PWHL data visualizations"
author: "Brittny Vongdara"
format: 
  revealjs:
    theme: solarized
    transition: slide
    background-transition: fade
footer: "RLadies 2024 IWD data visualization submission"
---

# International Women's day

```{r setup}
#| echo: false

# Install package ----
if (!requireNamespace('devtools', quietly = TRUE)){
  install.packages('devtools')
}
devtools::install_github(repo = "sportsdataverse/fastRhockey")

# Load libraries ----
library(fastRhockey)
library(tidyverse)
library(wordcloud2)
library(ggridges)
library(RColorBrewer)

# Get data ----
teams <- 
  pwhl_teams()

standing <- 
  pwhl_standings() %>% 
  mutate(across(c(games_played:penalty_kill_pct), as.numeric))

schedule <- 
  pwhl_schedule(season = 2023) %>% 
  select(-c(home_team_id, away_team_id, venue_url)) %>% 
  mutate(game_date = as.Date(strptime(sub(".*, ", "", game_date), "%b %d"))) %>% 
  filter(game_date <= Sys.Date())

stats_goalie <- 
  pwhl_stats(season = 2023) %>% 
  select(-minutes) %>% 
  mutate(across(c(games_played:so_save_percentage), as.numeric))

# stats_player <- 
  # do.call(rbind, lapply(c(unique(teams$team_label)), function(x) {
  #   pwhl_stats(season = 2023, position = "skater", team = x)
  # }))
  # pwhl_stats(position = "skater", team = "BOS", season = 2023)

players <- 
  do.call(rbind, lapply(c(unique(teams$team_label)), function(x) {
    pwhl_team_roster(season = 2023, team = x)
  }))

rm(teams)

w_v_l <-
  standing %>%
  select(team, wins, losses) %>%
  mutate(
    team = sub("PWHL ", "", team),
    losses = losses * -1) %>%
  pivot_longer(
    cols = c("wins", "losses"),
    names_to = "w_l",
    values_to = "score"
  )
```

International Women’s Day takes place on March 8th, 2024. The theme for this year is “Invest in women: Accelerate progress”.

More information can be found [here](https://www.internationalwomensday.com/).

## Investing in women

The data challenge put forth by the [RLadies Ottawa branch](https://rladies-ottawa.github.io/index.html) is to analyze investing in women.[^1]

[^1]: Find more information on the data challenge [here](https://rladies-ottawa.github.io/events/2024/iwd.html).

So what came to mind when I thought about investing in women?

## Professional Women's Hockey League (PWHL)

PWHL is a professional women's ice hockey league that began in January 2024. This league consists of 6 teams, 3 from Canada and the U.S. respectively:

- Montreal
- Ottawa
- Toronto
- Boston
- Minnesota
- New York

## What does this have to do with investing in women?

Typically women's sports don't have the same kind of reception as men's sports--which sucks a lot. i.e, women's sports don't get as much screen time[^3] and they don't get paid as much as their male counterparts[^4]

[^3]: According to a 2019 survey, coverage on female athletes on televised news was at 5.7% on SportsCenter from 2.0% in 2014. [Source: One and Done: The Long Eclipse of Women’s Televised Sports, 1989–2019 by Cheryl Cooky, et al](https://journals.sagepub.com/doi/full/10.1177/21674795211003524)

[^4]: A study by Royal Bank of Canada and Wasserman and The Collective found that, on average, male athletes earns 21 times more than professional female athletes. [Source: The Economy of Sports](https://static1.squarespace.com/static/63dd2a294ccbcf78ab818f33/t/656a2e3a29948f2d3d6b386e/1701457475152/The+Collective_RBC_Case+for+Womens+Sport_Pt+1_FINAL_ENG.pdf)

## Now the good news ... {.smaller}

However, things are starting to look up. In 2022, women's sports media coverage tripled (from 5.7% to 15%) from 2019.[^5]

[^5]: [Source: Media Coverage For Women’s Sports Has Nearly Tripled In Five Years, According To New Research by Lindsay Darvin](https://www.forbes.com/sites/lindseyedarvin/2023/10/31/media-coverage-for-womens-sports-has-nearly-tripled-in-five-years-according-to-new-research/?sh=15b7c1975ebb)

The PWHL is far from the first professional women's sport league but we're seeing a lot of positive reception from this emerging league, such as the sold out openers in Ottawa, Montreal, and Toronto.[^6] Not only that but the PWHL broke the all-time attendance record for women's hockey.[^7] News like this indicate the demand for women's sports.

[^6]: [Source: Canada's PWHL Teams Are Sold Out For Openers, Ottawa Smashes Record by Ian Kennedy](https://thehockeynews.com/womens/pwhl/canadas-pwhl-teams-are-sold-out-for-openers)

[^7]: [Source:Professional women’s hockey league breaks all-time attendance record for women’s hockey](https://www.thepwhl.com/en/professional-womens-hockey-league-breaks-attendance-record)

The PWHL is just one slice in the big sports picture but if we continue to give opportunities for professional women's sports league then we give them the opportunities to succeed and thrive.

In sum, when we **invest in women's professional sports we invest in women**!

# The data visualizations

So let's get onto some fun data visualizations about the PHWL.

## Horizontal bar chart

```{r horizontal_bar}
ggplot(schedule) +
  geom_histogram(aes(
    x = fct_rev(fct_infreq(venue)), 
    y = ..count..),
    fill = "#340789",
    stat = "count") +
  scale_y_continuous(breaks = seq(0, 12, 5)) +
  theme_classic() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.border = element_blank(),
    axis.ticks.x = element_blank(),
    panel.background = element_rect(fill = "#fdf6e3"),
    plot.background = element_rect(fill = "#fdf6e3", color = NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.background = element_rect(fill = "#fdf6e3"),
    legend.box.background = element_rect(fill = "#fdf6e3")
  ) +
  labs(
    title = "Frequency of games per venue",
    x = "Venue",     
    y = NULL) +
  coord_flip() 
```

## ggridges

```{r ggridges}
players %>%
  ggplot(aes(x = age, y = team, fill = team)) +
  geom_density_ridges(alpha = 0.6, stat = "binline", bins = 8) +
  theme_ridges() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.border = element_blank(),
    axis.ticks.x = element_blank(),
    panel.background = element_rect(fill = "#fdf6e3"),
    plot.background = element_rect(fill = "#fdf6e3", color = NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.background = element_rect(fill = "#fdf6e3"),
    legend.box.background = element_rect(fill = "#fdf6e3")
  ) +
  labs(
    title = "Distribution of ages by team",
    x = "Age",
    y = "Team"
  )
```

## Lollipop

```{r lollipop}
ggplot(stats_goalie,
       aes(x = fct_reorder(player_name, -desc(games_played)),
           y = games_played, colour = team)) +
  geom_segment(
    aes(x = player_name,
        xend = player_name,
        y = 0,
        yend = games_played), color = "grey") +
  geom_point(size = 4) +
  scale_color_manual(values = c("#154737", "#2E1A47", "#862633", "#00B2A9", "#A6192E", "#307FE2")) +
  theme_classic() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.border = element_blank(),
    axis.ticks.x = element_blank(),
    panel.background = element_rect(fill = "#fdf6e3"),
    plot.background = element_rect(fill = "#fdf6e3", color = NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.background = element_rect(fill = "#fdf6e3"),
    legend.box.background = element_rect(fill = "#fdf6e3")
  ) +
  labs(
    title = "Number of wins",
    x = "Goalie",
    y = ""
  ) +
  coord_flip()
```

## Mirrored histogram

```{r mirrored_histogram}
ggplot(w_v_l,
       aes(x = team,
           y = score,
           fill = w_l)) +
  geom_bar(stat = "identity", position = "identity") +
  ylim(-max(w_v_l$score), max(w_v_l$score)) +
  scale_y_continuous(breaks = seq(-max(w_v_l$score), max(w_v_l$score), 5)) +
  theme_classic() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.border = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_text(
      angle = 90,
      hjust = 1,
      vjust = 0.5,
      size = 15
    ),
    panel.background = element_rect(fill = "#fdf6e3"),
    plot.background = element_rect(fill = "#fdf6e3", color = NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.background = element_rect(fill = "#fdf6e3"),
    legend.box.background = element_rect(fill = "#fdf6e3")
  ) +
  labs(
    title = "Wins vs. losses",
    x = "Team",
    y = ""
  )
```

# Looking at the code

## Let's take my favourite visualization again

```{r ggridges2}
players %>%
  ggplot(aes(x = age, y = team, fill = team)) +
  geom_density_ridges(alpha = 0.6, stat = "binline", bins = 8) +
  theme_ridges() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.border = element_blank(),
    axis.ticks.x = element_blank(),
    panel.background = element_rect(fill = "#fdf6e3"),
    plot.background = element_rect(fill = "#fdf6e3", color = NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.background = element_rect(fill = "#fdf6e3"),
    legend.box.background = element_rect(fill = "#fdf6e3")
  ) +
  labs(
    title = "Distribution of ages by team",
    x = "Age",
    y = "Team"
  )
```

# Let's break it down

I used the `fastRhockey` package to return PWHL data on players from each team. Admittedly, the data behind this data visualization is kind of boring but it's a pretty plot. I found this chart [here](https://r-graph-gallery.com/294-basic-ridgeline-plot.html#shape).

This plot shows the age distribution by team using the `ggridges` package and the function `geom_density_ridges`.


## {auto-animate="true"}

Here are the building phases of this plot.

1. Setting up the shape

```r
players %>% # return PWHL player data
  ggplot(aes(x = age, y = team, fill = team)) + # X is age, and Y and fill are team
  geom_density_ridges(alpha = 0.6, stat = "binline", bins = 8)
```

## {auto-animate="true"}

2. Adding our details

```r
players %>%
  ggplot(aes(x = age, y = team, fill = team)) +
  geom_density_ridges(alpha = 0.6, stat = "binline", bins = 8) +
  theme_ridges() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.border = element_blank(),
    axis.ticks.x = element_blank(),
    panel.background = element_rect(fill = "#fdf6e3"),
    plot.background = element_rect(fill = "#fdf6e3", color = NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.background = element_rect(fill = "#fdf6e3"),
    legend.box.background = element_rect(fill = "#fdf6e3")
  )
```

## {auto-animate="true"}

3. Adding the labels

```r
players %>%
  ggplot(aes(x = age, y = team, fill = team)) +
  geom_density_ridges(alpha = 0.6, stat = "binline", bins = 8) +
  theme_ridges() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.border = element_blank(),
    axis.ticks.x = element_blank(),
    panel.background = element_rect(fill = "#fdf6e3"),
    plot.background = element_rect(fill = "#fdf6e3", color = NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.background = element_rect(fill = "#fdf6e3"),
    legend.box.background = element_rect(fill = "#fdf6e3")
  ) +
  labs(
    title = "Distribution of ages by team",
    x = "Age",
    y = "Team"
  )
```

# Let's look it again with the outputs

## Using a template

```{r}
#| echo: true
players %>% # return PWHL player data
  ggplot(aes(x = age, y = team, fill = team)) + # X is age, and Y and fill are team
  geom_density_ridges(alpha = 0.6, stat = "binline", bins = 20)
```

## Customizing the plot

This looked kind of weird so I ended up playing with the bin size and found that 8 was my Goldilocks number. 7 was too dense and 9 was too sparse.

```{r}
#| echo: true
players %>% # return PWHL player data
  ggplot(aes(x = age, y = team, fill = team)) + # X is age, and Y and fill are team
  geom_density_ridges(alpha = 0.6, stat = "binline", bins = 8)
```

## Nailing down the aesthetics and labels {.smaller}

It's the little things now. The plots stand out from the slides because of the white background so I want to make the background the same to have a seamless look. And we want to make sure that a read can understand what the plot is trying to convey so we want to appropriately name the labels.

```r
players %>%
  ggplot(aes(x = age, y = team, fill = team)) +
  geom_density_ridges(alpha = 0.6, stat = "binline", bins = 8) +
  theme_ridges() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.border = element_blank(),
    axis.ticks.x = element_blank(),
    panel.background = element_rect(fill = "#fdf6e3"),
    plot.background = element_rect(fill = "#fdf6e3", color = NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.background = element_rect(fill = "#fdf6e3"),
    legend.box.background = element_rect(fill = "#fdf6e3")
  ) +
  labs(
    title = "Distribution of ages by team",
    x = "Age",
    y = "Team"
  )
```

## Final output

```{r}
players %>%
  ggplot(aes(x = age, y = team, fill = team)) +
  geom_density_ridges(alpha = 0.6, stat = "binline", bins = 8) +
  theme_ridges() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.border = element_blank(),
    axis.ticks.x = element_blank(),
    panel.background = element_rect(fill = "#fdf6e3"),
    plot.background = element_rect(fill = "#fdf6e3", color = NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.background = element_rect(fill = "#fdf6e3"),
    legend.box.background = element_rect(fill = "#fdf6e3")
  ) +
  labs(
    title = "Distribution of ages by team",
    x = "Age",
    y = "Team"
  )
```

# The end

Thanks for looking at my plots and hopefully you're inspired to make your own pretty plots!

# Resources

- fastRhockey
  - Authors: Ben Howell and Saiem Gilani
  - Description: The SportsDataverse's R Package for Hockey Data
  - URL: [https://fastRhockey.sportsdataverse.org/](https://fastRhockey.sportsdataverse.org/)

- [Hex colour codes for PWHL teams](https://twitter.com/_MelissaBurgess/status/1724443118718640321)
